---
title: "Analysis of Malaria Data Set in R"
author: "Soong Zheng Hong Joel"
date: "10/11/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**LOAD PACKAGES**
```{r}
library(dplyr)
library(cowplot)
library(ggplot2)
```

**LOAD DATA**
```{r}
# Reading all 3 Malaria CSV files from Github URL: https://github.com/rfordatascience/tidytuesday/tree/master/data/2018/2018-11-13
# Malaria_deaths CSV file (Malaria deaths by country for all ages across the world and time)
data1 <- read.csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2018/2018-11-13/malaria_deaths.csv")

# Malaria_deaths_age CSV file (Malaria deaths by age across the world and time)
data2 <- read.csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2018/2018-11-13/malaria_deaths_age.csv")

# Malaria_inc CSV file (Malaria incidence by country for all ages across the world across time)
data3 <- read.csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2018/2018-11-13/malaria_inc.csv")
```

**DATA PREPARATION**
```{r}
# Checking for missing values in entries for all 3 data
is.null(data1)
is.null(data2)
is.null(data3)

# Drop redundant column 
# Data 1 to 3 Country Code is redundant as it serves no purpose for analysis
# Data 2 has an index column named "X" which is redundant
data1 <- data1[-c(2)]
data2 <- data2[-c(1,3)]
data3 <- data3[-c(2)]

# Changing the total deaths for data2 to be in per 100,000 to match data1 deaths measurement
data2$deaths <- (data2$deaths / 100000)

# Rename columns
colnames(data1) <- c("Country","Year", "Deaths_100000")
colnames(data2) <- c("Country", "Year", "Age_group", "Deaths_100000")
colnames(data3) <- c("Country", "Year", "Incidence_1000")

# Checking data type to ensure if variables are of correct data type
str(data1)
str(data2)
str(data3)

# Checking basic summary and exploration for all 3 data
summary(data1)
summary(data2)
summary(data3)
```

**DATA VISUALIZATION**
```{r}
# First visualization
# Group by Age Group with total deaths for data 1
by_age <- data2 %>% group_by(Age_group) %>% summarise_at(vars(Deaths_100000),funs(sum(.,na.rm=TRUE)))

# Plot Pie chart to see percentages of death for each age group
percentage <- round(by_age$Deaths_100000/sum(by_age$Deaths_100000)*100)
by_age$Age_group <- paste(by_age$Age_group, percentage) # Add percentages to labels
by_age$Age_group <- paste(by_age$Age_group,"%",sep="") # Add "%" to labels
pie(by_age$Deaths_100000, labels = by_age$Age_group, main="Total Number of Deaths/Age Group")
```
<br>
From this visualization, it is clearly shown that the age group with its highest percentage of deaths is **`r max(by_age$Age_group)`**.

```{r}
# Second Visualization
# Group by Country with total deaths 
by_country <- data1 %>% group_by(Country) %>% summarise_at(vars(Deaths_100000),funs(sum(.,na.rm=TRUE)))

# Retrieving top 5 countries with highest total number of deaths
by_country2 <- by_country[order(-by_country$Deaths_100000), ]
by_country_top5 <- head(by_country2, 5)

# Plot Histogram to see frequency of death for countries
hist(by_country$Deaths_100000,
     main = "Total Number of Deaths/Country",
     xlab = "Number of Deaths Per 100,000",
     ylab = "Number of Countries",
     col = "red")

# Plot Bar Chart to see Top 5 countries with highest total number of deaths
barplot(by_country_top5$Deaths_100000,names.arg=by_country_top5$Country,
        ylim = c(0,5000),
        xlab="Country",
        ylab="Number of Deaths Per 100,000",
        col="red",
        main="Top 5 Countries with Highest Total Number")
```
<br>
This visualization shows that majority of the countries have little to no deaths at all as the graph is highly skewed towards the value of 0.   
The top 5 countries that has the highest total number of deaths are **`r by_country_top5$Country`**.   
The highest number of cases across all countries stands at a value of **`r round((max(by_country$Deaths_100000)*100000))`**.
```{r}
# Third Visualization
# Group by Years with total deaths across all Countries
by_year <- data1 %>% group_by(Year) %>% summarise_at(vars(Deaths_100000),funs(sum(.,na.rm=TRUE)))

# Converting Year data type from INT to Date to plot time series
by_year$Year <- as.Date(paste(by_year$Year, 1, 1, sep = "-")) # beginning of year

# Group by Years with total Incidents across all Countries
by_yearInc <- data3 %>% group_by(Year) %>% summarise_at(vars(Incidence_1000),funs(sum(.,na.rm=TRUE)))

# Converting Year data type from INT to Date to plot time series
by_yearInc$Year <- as.Date(paste(by_yearInc$Year, 1, 1, sep = "-")) # beginning of year

# Plot Time series to see the trend of deaths over the years
plot1 <-  ggplot(data = by_year, aes(x = Year, y = Deaths_100000))+
          geom_line(color = "#00AFBB", size = 2)

# Adding title and editing axis name
plot1a <- plot1 + ggtitle("Total Number of Death") +
          xlab("Year") + ylab("Number of Deaths")

# Plot Time series to see the trend of Incidence over the years
plot2 <-  ggplot(data = by_yearInc, aes(x = Year, y = Incidence_1000))+
  geom_line(color = "#FC4E07", size = 2)

# Adding title and editing axis name
plot2a <- plot2 + ggtitle("Total Number of Incidence") +
          xlab("Year") + ylab("Number of Incidence")

# Put graphs in 2 rows and 1 column to see both time series plot in a comparison view
plot_grid(plot1a, plot2a, 
          labels = c("A", "B"),
          ncol = 1, nrow = 2)
```
<br>
This visualization shows that although the number of deaths was on an increasing trend from 1990s to 2000, the number of deaths and incidences from 2000 henceforth is in a downward trend.  
Steady decline in trend for both plots suggests that:  
1) There is a positive correlation between deaths and incidences as one declines, the other will follow suit  
2) Malaria is being curbed well across these countries as deaths and incidences are declining over time.